Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-15 EBDVF-2024/paper.tex
Add files via upload

\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDSA RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the vocabulary hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.
\end{abstract}

@@

In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
the architecture of dataspaces in the reference architecture model
IDS-RAM (current version 4). One of the defined architectural
components, the Vocabulary Hub, has the role of providing vocabularies
to annotate and describe data assets and services. These vocabularies
form a common language to enable semantic interoperability among the
dataspace participants. The Vocabulary Hub is responsible for resolving
annotations of data assets and for providing details in the form of
standardized semantic descriptions. It supports the use of Semantic Web
standards for defining and representing RDF vocabularies and ontologies
and thereby provides semantic interoperability in a standardized and
machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not
only those designed in accordance with IDS-RAM. Other reference
architecture models, such as GAIA-X architecture document 22.10 Release
\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
Architecture Model for the Industrial Data Space
\citep{ReferenceArhictectureModel2017} also describe components with a
similar role - the Federated Catalogue and Broker Service, respectively.
Therefore, our approach designed around the IDS-RAM is adaptable and
applicable to data spaces designed in accordance with other reference
architecture models.

@@

The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic
building block to achieve semantic interoperability in dataspaces by
provisioning common vocabularies. Those vocabularies are to be expressed
in RDF and by syntactically compatible with OWL to allow for expressive
semantics. The IDS-RAM does not require, but nonetheless allows for,
extending the function of the Vocabulary hub towards providing ontology
mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
dataspace connectors to automatically convert data between user-specific
and standard data formats. Such automatic conversions are to be
performed by Data Tranformation Apps by either the Data Consumers, the
Data Providers, or both.

@@

\begin{itemize}
\tightlist
\item
  Bringing it into actual dataspaces in practice, representing and
  integrating datasets from different dataspace participants.
\item
  Implementing use cases for richer discoverability, harmonized querying
  and support for different content types of structured and unstructured
  data.
\item
  Explore how ML can benefit regarding quality in practice when
  providing consolidated and cleaned data via dataspaces.
\item
  Discuss how we can extend the IDS-RAM with services to improve the
  support for semantic interoperability provided by our solution.
\end{itemize}

>>>>

\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDS RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the vocabulary hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.
\end{abstract}

@@

In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
the architecture of dataspaces in the reference architecture model IDS
RAM (current version 4). One of the defined architectural components,
the Vocabulary Hub, has the role of providing vocabularies to annotate
and describe data assets and services. These vocabularies form a common
language to enable semantic interoperability among the dataspace
participants. The Vocabulary Hub is responsible for resolving
annotations of data assets and for providing details in the form of
standardized semantic descriptions. It supports the use of Semantic Web
standards for defining and representing RDF vocabularies and ontologies
and thereby provides semantic interoperability in a standardized and
machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not
only those designed in accordance with IDS RAM. Other reference
architecture models, such as GAIA-X architecture document 22.10 Release
\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
Architecture Model for the Industrial Data Space
\citep{ReferenceArhictectureModel2017} also describe components with a
similar role - the Federated Catalogue and Broker Service, respectively.
Therefore, our approach designed around the IDS RAM is adaptable and
applicable to data spaces designed in accordance with other reference
architecture models.

@@

The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic
building block to achieve semantic interoperability in dataspaces by
provisioning common vocabularies. Those vocabularies are to be expressed
in RDF and by syntactically compatible with OWL to allow for expressive
semantics. The IDS RAM does not require, but nonetheless allows for,
extending the function of the Vocabulary hub towards providing ontology
mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
dataspace connectors to automatically convert data between user-specific
and standard data formats. Such automatic conversions are to be
performed by Data Tranformation Apps by either the Data Consumers, the
Data Providers, or both.

@@

\begin{itemize}
\tightlist
\item
  Bringing it into actual dataspaces in practice, representing and
  integrating datasets from different dataspace participants.
\item
  Implementing use cases for richer discoverability, harmonized querying
  and support for different content types of structured and unstructured
  data.
\item
  Explore how ML can benefit regarding quality in practice when
  providing consolidated and cleaned data via dataspaces.
\item
  Discuss how we can extend the IDS RAM with services to improve the
  support for semantic interoperability provided by our solution.
\end{itemize}
